---
title: "COVID19_steps"
author: ""
date: "2024-04-16"
output:
  pdf_document: default
  html_document: default
---
Import libraries
```{r import}
library(tidyverse)
```

Read in the data from the csv file.
```{r get_jhu_data}
## Get current data in the four files
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

file_names <- c("time_series_covid19_confirmed_US.csv",  "time_series_covid19_confirmed_global.csv", "time_series_covid19_deaths_US.csv",  "time_series_covid19_deaths_global.csv")

urls <- str_c(url_in, file_names)
```

Read in the data to see what we have.

```{r import_data, message = FALSE}
global_cases <- read_csv(urls[2])
global_deaths <- read_csv(urls[4])
US_cases <- read_csv(urls[1])
US_deaths <- read_csv(urls[3])

```

Merge date columns and case numbers for global_cases.

```{r tidy_global_cases}
global_cases <- global_cases %>% pivot_longer(
  cols = matches("\\d{1,2}/\\d{1,2}/\\d{2}"),
  names_to = "date",
  values_to = "cases")
```

Remove unneeded columns for global_cases.

```{r remove_global_cases}
global_cases_clean <- global_cases %>%
  select(-c(Lat, Long))
```

Merge date columns and case numbers for global_deaths.

```{r tidy_global_deaths}
global_deaths <- global_deaths %>% pivot_longer(
  cols = matches("\\d{1,2}/\\d{1,2}/\\d{2}"),
  names_to = "date",
  values_to = "deaths"
)

```

Remove unneeded columns for global_deaths.

```{r remove_global_deaths}
global_deaths_clean <- global_deaths %>% 
  select(-c(Lat, Long))
```

Merge

```{r merge_cases_and_deaths}
global <- global_cases_clean %>% full_join(global_deaths_clean)
```
Fix Dates

```{r date_correct}
global <- global %>% mutate(date = myd(date))
```

Process US_cases

```{r tidy_us_cases}
US_cases <- US_cases %>% pivot_longer(
  cols = -(UID:Combined_Key),
  names_to = "date",
  values_to = "cases") %>%
  select(Admin2:cases) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))
```

```{r tidy_us_deaths}
US_deaths <- US_deaths %>% pivot_longer(
  cols = -(UID:Population),
  names_to = "date",
  values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))
```

Join them.

```{r us_join}
US <- US_cases %>%
  full_join(US_deaths)
```

```{r fix_global_names}
global <- rename(global, Province_State = `Province/State`,
                 Country_Region = `Country/Region`)
```


```{r preparing_US_and_global}
global <- global %>%
  unite("Combined_Key",
        c(Province_State, Country_Region),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE)
```

```{r read_in_extra}
uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"

uid <- read_csv(uid_lookup_url) %>%
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))
```

```{r}
global <- global %>% 
  left_join(uid, by = c("Province_State", "Country_Region")) %>%
  select(-c(UID, FIPS)) %>%
  select(Province_State, Country_Region, date, cases, deaths, Population, Combined_Key)
```

```{r}
US_by_state <- US %>% group_by(Province_State, Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths), 
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths *1000000 / Population) %>%
  select(Province_State, Country_Region, date, cases, deaths, deaths_per_mill, Population) %>% ungroup()
```
```{r}
US_totals <- US_by_state %>% group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths), 
            Population = sum(Population)) %>%
  mutate(deaths_per_mill = deaths *1000000 / Population) %>%
  select(Country_Region, date, cases, deaths, deaths_per_mill, Population) %>% ungroup()
```
```{r filter_US_totals}
US_totals %>% 
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position="bottom",
        axis.text.x = element_text(angle = 90)) +
    labs(title = "COVID-19 in US", y= NULL)
  
```
```{r state_only}
state <- "New York"
US_by_state %>%
  filter(Province_State == state) %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position="bottom",
        axis.text.x = element_text(angle = 90)) +
    labs(title = str_c("COVID-19 in ", state), y= NULL)
```
```{r transform_again}
US_by_state <- US_by_state %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))

US_totals <- US_totals %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))
```

```{r plot_new_cases_new_deaths}
US_totals %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme(legend.position="bottom",
        axis.text.x = element_text(angle = 90)) +
    labs(title = "COVID-19 in US", y= NULL)
```
```{r}
US_state_totals <- US_by_state %>%
  group_by(Province_State) %>%
  summarize(deaths = max(deaths), cases = max(cases),
            population = max(Population),
            cases_per_thou = 1000* cases / population,
            deaths_per_thou = 1000* deaths / population) %>%
  filter(cases > 0, population > 0)
```

```{r slice_min}
US_state_totals %>%
  slice_min(deaths_per_thou, n=10) %>%
  select(deaths_per_thou, cases_per_thou, everything())
```
```{r slice_max}
US_state_totals %>%
  slice_max(deaths_per_thou, n=10) %>%
  select(deaths_per_thou, cases_per_thou, everything())
```
```{r models}
mod <- lm(deaths_per_thou ~ cases_per_thou, data = US_state_totals)
summary(mod)
```
```{r smallest}
US_state_totals %>% slice_min(cases_per_thou)
```
```{r}
US_state_totals %>% slice_max(cases_per_thou)
```
```{r}
US_state_totals %>% mutate(pred = predict(mod))
```
```{r}
US_tot_w_pred <- US_state_totals %>% mutate(pred = predict(mod))
```

```{r}
US_tot_w_pred %>% ggplot() +
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") +
  geom_point(aes(x = cases_per_thou, y = pred), color = "red")
```
Biases:
I could read into this data more than I am, but I've grounded myself in it because there are underlying factors that this visualization doesn't account for and all I can say conclusively is that case numbers and deaths positively correlate.
